---
title: "Analyse Experiments"
author: "Anthony Stephenson"
date: "1/18/2021"
output: pdf_document
---

```{r import4, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(fs)
library(purrr)
library(lhc)
```

# Analyse Experiments
As mentioned previously, we ran a series of experiments running a logistic regression model with varying model parameters. The experiments were run in a separate script `RunModels.R`, and the output of the experiments are saved in the `output` folder of the git repository. 

## Import results
```{r import}
filepath <- path_join(c(dirname(getwd()), "/output/results_6.csv"))
exp_data <- read.csv(filepath) %>% select(-X)
```

Experiments were run over all combinations of the following parameters:
```{r}
#the number of rbf features
(n_rbfs <- unique(exp_data$n_rbf))

#the l2 regularisation parameter
(lambdas <- unique(exp_data$lambda))

#and the degree of polynomial transform (with no interaction terms)
(poly_orders <- unique(exp_data$poly))
```

Since we want a model with high performance but low variance, we measured the median absolute deviation (mad) of the AMS and AUC metrics over the k-folds. Now we can look to maximise the mean(AMS)/mad(AMS).

```{r}
# scaling metrics by MAD(metric) to get a handle on variation
exp_data$scaled.auc <- exp_data[,"auc"]/exp_data[,"mad.auc."]
exp_data$scaled.ams <- exp_data[,"ams"]/exp_data[,"mad.ams."]
```


## Plot metrics
```{r, fig.width=5, fig.height=3.5}
#convert columns to factors for boxplots
exp_data$lambda <- signif(exp_data$lambda,3)
cols <- c("G", "lambda", "poly", "n_rbf")
exp_data[cols] <- lapply(exp_data[,cols], factor) 

(p1 <- exp_data %>%
  ggplot(aes(x=n_rbf, y=ams, colour=G)) +
    geom_boxplot() +
    theme_minimal() +
    labs(x="Number of RBF features", y="AMS",
         title="AMS of models with n RBF features",
         colour="Jet Group"))

(p2 <- exp_data %>%
  ggplot(aes(x=poly, y=ams, colour=G)) +
    geom_boxplot() + 
    theme_minimal() +
    labs(x="Order of polynomial transform", y="AMS",
         title="AMS of models with b degree polynomal transform",
         colour="Jet Group"))

(p3 <- exp_data %>%
  ggplot(aes(x=lambda, y=ams, colour=G)) +
    geom_boxplot() + 
    theme_minimal() +
    labs(x="Strength of l2 regularisation", y="AMS",
         title="AMS of models with lambda regularisation",
         colour="Jet Group"))
```

## Record results
```{r, include=F, message=F, echo=F}
pdf("../doc/figs/Experiments_nrbf.pdf", width=5, height=4)
p1
dev.off()
pdf("../doc/figs/Experiments_poly.pdf", width=5, height=4)
p2
dev.off()
pdf("../doc/figs/Experiments_lambda.pdf", width=5, height=4)
p1
dev.off()
```

Print results as a LaTeX table that can be imported into a report easily. Not too surprisingly we see different models appear at the top of our list depending on how we decide to choose the best model. Since our goal is ultimately to maximise AMS, it makes sense that this (rather than AUC) should be the metric we choose, despite the higher variance it displays. To try and mitigate this variance, we can  choose to sort by a scaled version of AMS, where we take the mean absolute deviation across the folds and scale by that, to try and optimise for the best result with the least variation and therefore hopefully the best generalisation.

The top models appears to have the following set of parameters: $n_{rbf}=2$, $\lambda = 1e-4$ and the inclusion of up to 3rd order polynomial terms.

```{r latex output, warning=FALSE, message=FALSE}
library(Hmisc)
#For each group find the experiment with the highest ams
output <- NULL
for(i in 1:3){
  # exp_sub <- filter(exp_data, G==i) %>%
  #   arrange(desc(scaled.ams), desc(scaled.auc))
  
  exp_sub <- filter(exp_data, G==i) %>%
    arrange(desc(ams))
  
  output <- rbind(output, exp_sub[1,])
}

print(output)

# make column names latex friendly
colnames(output) <- sub("(\\w+)\\.(\\w+)\\.?", "\\$\\1_\\{\\2\\}\\$", colnames(output))
colnames(output) <- sub("n\\_rbf", "\\$n_{rbf}\\$", colnames(output))

output[7:12] <- round(output[7:12], 3)

# Generate LaTeX table
latex(output, file=path_join(c(dirname(getwd()), "doc/results_table_groups.tex")), caption="Results table of top experiment for each of the groups", label="table:results")
```
\input{../doc/results_table}

